---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fabletools
<!-- badges: start -->
[![R build status](https://github.com/tidyverts/fabletools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tidyverts/fabletools/actions/workflows/R-CMD-check.yaml)
[![Code test coverage](https://codecov.io/gh/tidyverts/fabletools/branch/master/graph/badge.svg)](https://app.codecov.io/gh/tidyverts/fabletools?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![CRAN status](https://www.r-pkg.org/badges/version/fabletools)](https://CRAN.R-project.org/package=fabletools)
<!-- badges: end -->

The R package *fabletools* provides tools for building modelling packages, with a focus on time series forecasting. This package allows package developers to extend  *fable* with additional models, without needing to depend on the models supported by *fable*.

## Installation

You could install the **stable** version on [CRAN](https://cran.r-project.org/package=fabletools):

```{r, eval = FALSE}
install.packages("fabletools")
```

You can install the **development** version from
[GitHub](https://github.com/tidyverts/fabletools):

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("tidyverts/fabletools")
```
